# Working with data

Every analysis of data starts with preprocessing and parsing. The following chapter covers some packages that simplify working with data.

## tidyverse

```{r}
library(tidyverse)
data(iris)
head(iris)

dplyr::group_by(iris, Species) %>%
  dplyr::summarize(Petal.Length = mean(Petal.Length),
            Sepal.Length = mean(Sepal.Length)) %>%
  dplyr::mutate(Species = toupper(Species)) %>%
  head()

tidyr::gather(iris, Col, Val, -Species) %>%
  head()
```

TODO some examples

## data.table

`data.table` is a fast implementation of R's classical `data.frame`. I hardly ever use data frame any more, and if so only, because it seems to work nicer with `dplyr` and `tidyr`. However, by using `dtplyr` this isn't much of a problem, really.

```{r, fig.width=2, fig.height=2, fig.align = 'center'}
library(data.table)  
library(dplyr)
library(dtplyr)
library(grid)
library(gridExtra)

n    <- 1000
rn   <- stats::rnorm(n)
ltrs <- base::sample(letters[1:5], n, replace=TRUE)
dt   <- data.table::data.table(X=rn, Y=ltrs)
df   <- base::data.frame(X=rn, Y=ltrs)

dt[, .SD[sample(.N, 1)], by=c("Y")] %>%
  tableGrob(rows=NULL) %>%
  grid.arrange
```

In the end it depends what style you prefer. I usually go with `data.table` alone without needing the `dplyr/dtplyr` dependency. However, the latter is usually more readable. For large data, the fastest solution is probably preferable.

```{r}
  dt.only   <- function() dt[, .SD[sample(.N, 1)], by=c("Y")]
  dt.dtplyr <- function() dt %>% dplyr::group_by(Y) %>% dplyr::sample_n(1)
  df.dplyr  <- function() df %>% dplyr::group_by(Y) %>% dplyr::sample_n(1)

  microbenchmark::microbenchmark(
    dt.only(),
    dt.dtplyr(),
    df.dplyr()
  )
```

## Data structures

If you have a background in computer science you may wonder, that R does not have support for advanced data structures such as Fibonacci heaps or hashmaps.
`datastructures` tries to solve this. It uses Rcpp modules to export Boost data structures to R:

```{r}
  library(datastructures)
  q <- datastructures::fibonacci_heap("integer")
  q[1:3] <- list(rnorm(3), 2, rnorm(4))
  datastructures::pop(q)
  datastructures::pop(q)
  datastructures::pop(q)
  datastructures::pop(q)
```

TODO examples

## Databases

TODO neo4j, maria, sqltie, mongodb, dbplyr
